---
title: "W7 assignment"
author: "Anna Stuckert"
date: "10/23/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Loading packages
library(tidyverse)
library(gapminder)

df <- gapminder
```


	1. Define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.
	
```{r}

calcGDP <- function(df) { #takes the dataframe of choice as input
  if (!is.numeric(df$gdpPercap)) {
    stop("gdpPercap must be a numeric vector.")
  }
  gdp <- df$pop * df$gdpPercap #multiplies the population number by the gross doemstic product per citizen
  return(gdp) #returns a vector of the same length as your dataframe, containing the GDP for each observation.
}

#using the function to add a column to the gapmidner dataframe with GDP
df$GDP <- calcGDP(df)

# filtering the DF to only include Denmark from the years of interest
df_DK <- filter(df, country == "Denmark", year == 1967|year == 1977|year == 1987|year == 1997| year == 2007| year == 2017)

#2017 isn't added to the dataframe, as the variable seemingly doesn't exist in the original dataframe

head(df_DK) #viewing the newly created dataframe including danish GDP values from the years of interest
```

	
	2. Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘B’ , and print out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70.
	
```{r}
#Convert country variables to characters
df$country <- as.character(df$country)

#as a single function

for (i in df$lifeExp){
  for (country in df$country) {
    if(startsWith(country, 'B')) {
      print("country starts with B")
    } else { 
        print("country doesn't start with B")
      }
  }
  if (i < 50) {
    print("Life expectancy is below 50")
  } else if (between(i, 50,70)) {
    print("Life expectancy is between 50 and 70")
  } else {
    print("Life expectancy is greater than 70")
    }
}
  

#as 2 seperate functions, one for whether country starts with B and one for life expectancy levels
  for (country in df$country) {
    if(startsWith(country, 'B')) {
      print("country starts with B")
    } else { 
        print("country doesn't start with B")
      }
  }

for (i in df$lifeExp){
  if (i < 50) {
    print("Life expectancy is below 50")
  } else if (between(i, 50,70)) {
    print("Life expectancy is between 50 and 70")
  } else {
    print("Life expectancy is greater than 70")
    }
}

```
	
