---
title: "Assignment w8"
author: "Anna Stuckert"
date: "11/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Loading packages
library(rvest)
library(janitor)
library(tidyverse)
```

1) adapt the webscraping example to scrape homicide data from FBI site and produce a meaningful report on how homicide trends evolve around US in relation to this urban unrest
```{r}
# scrape the website
url_FBI <- read_html("https://en.wikipedia.org/wiki/List_of_U.S._states_by_homicide_rate") #Reading data from html format from website

table_FBI <- html_table(url_FBI, fill = TRUE) #converting information into table format

FBI_df <- table_FBI[[1]] #selecting specific table of interest from the list of tables.

FBI_df #viewing table
```



2) use the rvest library to scrape data of your interest (football statistics in Wikipedia? global population by country in https://www.worldometers.info/world-population/population-by-country/ )

```{r}
url_population <- read_html("https://www.worldometers.info/world-population/population-by-country/")

table_population <- html_table(url_population)

population_df <- table_population[[1]]

population_df <- clean_names(population_df)
```


#To practice scraping, I performed the exercise again, scraping data related to the corona virus
```{r}
url_corona <- read_html("https://www.bbc.com/news/world-51235105")  #Reading data from html format from website

table_corona <- html_table(url_corona)  #converting information into table format

#if you have a long list of tables you can loop over them to find the one of interest quickly
for (i in table_corona){
  print(head(i))
}

#turn the desired table into a dataframe
corona_df <- table_corona[[1]]

#clean the names using janitor
corona_df <- clean_names(corona_df)


```

3) produce data visualisations that shed light on another interesting aspect of the police killing data

```{r}

#This is using the code from the homicide.Rmd provided by Adela earlier in the course

homicide <- FBI_df %>% 
  pivot_longer(-State, names_to = "year", values_to = "rate100k")
homicide$year <- as.numeric(homicide$year)  # to stop ggplot from acting up over character column
homicide

library(ggrepel)

#plotting the data for useful insights.
ggplot(homicide,
       aes(year,
           rate100k*10,
           colour = State)) +
  geom_line() +
	expand_limits(x=c(1994, 2022)) +
  geom_text_repel(data = homicide[homicide$rate100k > 7.5 & homicide$year == 2018,],
           aes(label = State,
               x = 2018.1, 
               y = rate100k*10, 
               colour = State, 
               hjust = -.01)) +
  theme_minimal()  +
  ylab("Number of people killed per 1,000,000") +
  xlab("Year") +
  theme(legend.position="none")

#We can also extract information related to each year and take the means to elucidate informaiton on which year had the highest crime rate
FBI_df <- FBI_df %>% select(-"State")
colMeans(FBI_df)

```

Meaningful report of the data:
Seemingly there is a peak in homicides at around 1996, then a donwward trend in homicides (except Louisiana), with a peak again at 2017 (precise year derived from colmun means using colMeans)


I didn't really know what questions to ask with the homicie data since Adela's plot provided in the homicides.rmd explained a lot, but instead I plotted my scraped corona-data from BBC
```{r}

#checking structure of data
str(corona_df)

#converting case count into numerical
corona_df$total_cases <- as.numeric(corona_df$total_cases)

#Research question: Is there a relationship between number of cases and death rate?
ggplot(corona_df, aes(total_cases, death_rate)) +
	geom_point() + geom_smooth(method = "lm")

# from the plot it would seem like there is a higher deathrate (not total death count, but rate) for higher count of infected - this is however likely due to a few values

#However, when looking at the dataframe, it is evident the table is not correct, as some countries have a deathrate of 0, even though they have reported deaths (e.g. Vietnam and Tanzania)
#Also it is possible some countries (possibly under dictatorship maybe???) do not report deaths (which is about 167 of them, bastards), which gives an incorrect impression
#Overall it's a pretty bad and uninformative dataset

arrange(corona_df, (death_rate))


```

